---
title: "Zebrafish MSigDB Gene Set Enrichment Testing"
author: "Nhi Hin"
date: "2020-04-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r load-packages, include=FALSE}
library(GSEABase)
library(dplyr)
library(readr)
library(magrittr)
library(tibble)
library(reshape2)
library(fgsea)
library(ggplot2)
library(org.Hs.eg.db)
library(org.Dr.eg.db)
library(biomaRt)
library(limma)
library(here)
library(stringr)
library(openxlsx)
library(pheatmap)
```


## 0. Introduction

- In this analysis, we will perform gene set enrichment testing on the 
fAD-like zebrafish dataset using the following gene sets:

  - MSigDB **Hallmark** gene set collection 
  - IRE gene sets defined in `ire-genes.Rmd`

- Previously in `de-analysis.Rmd`, we performed differential gene expression 
analysis using *limma*. 

- To perform gene set enrichment testing, we will use a combined method 
involving three different gene set enrichment testing methods (*fry*, 
*camera*, and *fgsea*). See `combinedGSEA.R` for details on how this is 
done. The function `combinedGSEA.R` requires some objects from the *limma* 
analysis (contrasts, design, voomData) to be imported. 

- The gene sets on MSigDB contain human genes. To use these as part of an 
analysis with zebrafish data, the homologous zebrafish gene IDs first 
need to be swapped in. 


## 1. Import .gmt files from MSigDB

- The gene set collections from MSigDB were downloaded from 
[here](https://www.gsea-msigdb.org/gsea/downloads.jsp) as .gmt files 
containing human entrezgene IDs. 

- Below, we import them in and convert from a list of lists to a 
data.frame. 

```{r}
genesetsDir <- here("data", "gsea")

h <- getGmt(file.path(genesetsDir, 
                      "h.all.v6.0.entrez.gmt"))    # Hallmark gene sets
c1 <- getGmt(file.path(genesetsDir,
                       "c1.all.v6.0.entrez.gmt"))  # positional gene sets
c2 <- getGmt(file.path(genesetsDir, 
                       "c2.all.v6.0.entrez.gmt"))  # curated gene sets
c3 <- getGmt(file.path(genesetsDir,
                       "c3.all.v6.0.entrez.gmt"))  # motif gene sets
c4 <- getGmt(file.path(genesetsDir, 
                       "c4.all.v6.0.entrez.gmt"))  # computational gene sets
c5 <- getGmt(file.path(genesetsDir, 
                       "c5.all.v6.0.entrez.gmt"))  # GO gene sets
c6 <- getGmt(file.path(genesetsDir,
                       "c6.all.v6.0.entrez.gmt"))  # oncogenic gene sets
c7 <- getGmt(file.path(genesetsDir,
                       "c7.all.v6.0.entrez.gmt"))  # immunologicgene sets

# Convert each gene sets to list where the name of each list is the gene set
# name and the list items are the entrezgenes. 
h_l <- geneIds(h) %>% as.list
c1_l <- geneIds(c1) %>% as.list
c2_l <- geneIds(c2) %>% as.list
c3_l <- geneIds(c3) %>% as.list
c4_l <- geneIds(c4) %>% as.list
c5_l <- geneIds(c5) %>% as.list
c6_l <- geneIds(c6) %>% as.list
c7_l <- geneIds(c7) %>% as.list

# Bind the list of gene sets so that each list becomes a data.frame.
h_df <- lapply(h_l, cbind)
c1_df <- lapply(c1_l, cbind)
c2_df <- lapply(c2_l, cbind)
c3_df <- lapply(c3_l, cbind)
c4_df <- lapply(c4_l, cbind)
c5_df <- lapply(c5_l, cbind)
c6_df <- lapply(c6_l, cbind)
c7_df <- lapply(c7_l, cbind)
```


## 2. Convert human entrezgenes in MSigDB gene sets to zebrafish ensembl IDs

- The IDs in the MSigDB gene sets are human entrezgene IDs which need to be 
converted into zebrafish Ensembl gene IDs. 

- We will prepare a data.frame `zebAndHumanEnsGenes` to map between the IDs.

```{r}
# Human entrez and ensembl IDs:
humanEntrezEns <- org.Hs.egENSEMBL %>% 
  as.data.frame %>%
  set_colnames(c("human_entrezgene", "human_ensembl"))

# Zebrafish ensembl IDs:
zebEns <- org.Dr.egENSEMBL %>%
  as.data.frame %>%
  set_colnames(c("zeb_entrezgene", "zeb_ensembl"))

# Create a data.frame to map between human entrezgenes & zebrafish ensembl IDs.
# BioMart only includes homolog mappings for ensembl IDs which is why we need to 
# retrieve human ensembl IDs, then join to the humanEntrezEns data.frame,
# in order to get the desired human entrezgenes to zebrafish ensembl ID mapping. 
# Note: Sometimes biomart doesn't work so I'm saving the mart out
# zebMart <- useMart("ensembl", "drerio_gene_ensembl")
zebMart <- readRDS(here("data", "zebrafish", "zebMart.rds"))
getFromBiomart <- c("ensembl_gene_id", "hsapiens_homolog_ensembl_gene")

zebAndHumanEnsGenes <- getBM(getFromBiomart, 
                             values = unique(zebEns$zeb_ensembl), 
                             mart = zebMart) %>%
  set_colnames(c("zebrafish_ensembl", "human_ensembl")) %>%
  left_join(humanEntrezEns, by = "human_ensembl") %>%
  dplyr::select(-human_ensembl) %>% 
  dplyr::filter(complete.cases(.))
```

- This mapping in `zebAndHumanEnsGenes` contains 
`r (zebAndHumanEnsGenes %>% lapply(function(x){x%>%unique%>%length}))[[1]]` 
zebrafish Ensembl IDs to 
`r (zebAndHumanEnsGenes %>% lapply(function(x){x%>%unique%>%length}))[[2]]` 
human entrezgene IDs. 

- The following function is applied to the gene sets to map the IDs in 
the gene sets from human entrezgene to zebrafish ensembl. 

```{r eval=FALSE}
mapHumanGS2Zebrafish <- function(x) {
  x %>% 
    as.data.frame %>% 
    set_colnames("human_entrezgene") %>%
    left_join(zebAndHumanEnsGenes, by = "human_entrezgene") %>% 
    dplyr::filter(complete.cases(.)) %>%
    dplyr::select(-human_entrezgene) %>%
    as.list %>%
    unname %>%
    .[[1]] %>%
    unique
}

h_mapped <- lapply(h_df, mapHumanGS2Zebrafish)
c1_mapped <- lapply(c1_df, mapHumanGS2Zebrafish)
c2_mapped <- lapply(c2_df, mapHumanGS2Zebrafish) 
c3_mapped <- lapply(c3_df, mapHumanGS2Zebrafish)
c4_mapped <- lapply(c4_df, mapHumanGS2Zebrafish)
c5_mapped <- lapply(c5_df, mapHumanGS2Zebrafish)
c6_mapped <- lapply(c6_df, mapHumanGS2Zebrafish)
c7_mapped <- lapply(c7_df, mapHumanGS2Zebrafish)
```

- Because the mapping takes a while to run, I've saved the mapped 
genesets as R objects, which can be imported as below:

```{r}
zebGenesets <- file.path(genesetsDir, "zebrafish")

h_mapped <- readRDS(file.path(zebGenesets, "ens_h_mapped.rds"))
c1_mapped <- readRDS(file.path(zebGenesets, "ens_c1_mapped.rds"))
c2_mapped <- readRDS(file.path(zebGenesets, "ens_c2_mapped.rds"))
c3_mapped <- readRDS(file.path(zebGenesets, "ens_c3_mapped.rds"))
c4_mapped <- readRDS(file.path(zebGenesets, "ens_c4_mapped.rds"))
c5_mapped <- readRDS(file.path(zebGenesets, "ens_c5_mapped.rds"))
c6_mapped <- readRDS(file.path(zebGenesets, "ens_c6_mapped.rds"))
c7_mapped <- readRDS(file.path(zebGenesets, "ens_c7_mapped.rds"))
```

## 3. Build indexes for MSigDB gene sets

- We now need to build an index for each collection of gene sets. 
The index is basically the ensembl IDs of the gene sets to the 
corresponding row number in the voom object previously generated in the 
differential gene expression analysis. 

- First we import the voom object:

```{r}
v <- readRDS(here("data", "zebrafish", "voomData_g.rds"))
```

- Build the indexes for each gene set collection: 

```{r}
buildGenesetIndex <- function(x, voomObj = v){
  limma::ids2indices(x, rownames(voomObj))
}

h_idx <- buildGenesetIndex(h_mapped)
c1_idx <- buildGenesetIndex(c1_mapped)
c2_idx <- buildGenesetIndex(c2_mapped)
c3_idx <- buildGenesetIndex(c3_mapped)
c4_idx <- buildGenesetIndex(c4_mapped)
c5_idx <- buildGenesetIndex(c5_mapped)
c6_idx <- buildGenesetIndex(c6_mapped)
c7_idx <- buildGenesetIndex(c7_mapped)
```

- The C2 collection contains all gene sets based upon KEGG pathways. 
We will subset this out as well in case we want to do any enrichment 
analyses specifically on KEGG gene sets. 

```{r}
kegg_idx <- c2_idx[(names(c2_idx) %>% str_detect("KEGG"))]
head(kegg_idx)
```

## 4. Build indexes for IRE gene sets

- The IRE gene sets were previously defined in `ire-zebrafish-genesets.Rmd`. 

- Here we will import the saved R objects of these gene sets back in 
and build the indexes. 

```{r}
ireGenes <- readRDS(here("data","ire","zebrafish","ireGenes.rds"))

ireGenes_idx <- buildGenesetIndex(ireGenes)
str(ireGenes_idx)
```


## 5. Perform Gene Set Testing

- The next step is to perform gene set testing using the MSigDB gene sets 
and the IRE gene sets. 

- We will need a few additional objects (design and contrast matrices)
 from the analysis with *limma* in `de-analysis.Rmd`. 

```{r}
design <- readRDS(here("data", "zebrafish", "design_g.rds"))
contrasts <- readRDS(here("data", "zebrafish", "contrasts_g.rds"))
```

- The gene set testing will be performed with *fry*, *camera* and *fgsea* 
with the raw p-values from this method being combined with Wilkinson's method. 
This is included in the `combinedGSEA` function which we will import below. 

```{r}
source(here("code","combinedGSEA.R"))
```

- The gene set testing will be performed with the following gene set 
collections:

    - **IRE gene sets** defined earlier. 
    - **KEGG gene sets** based on pathways from the KEGG database. 
    - **Hallmark gene sets** non-redundant gene sets representing 
    diverse biological / cellular pathways or activities. 

```{r eval=FALSE}
gseaResults_ire <- combinedGSEA(v, ireGenes_idx, design, contrasts)
gseaResults_kegg <- combinedGSEA(v, kegg_idx, design, contrasts)
gseaResults_h <- combinedGSEA(v, h_idx, design, contrasts)
```
```{r include=FALSE}
gseaResults_ire<-readRDS(here("data", "zebrafish", 
                                 "gsea", "gseaResults_ire.rds"))
gseaResults_kegg<-readRDS(here("data", "zebrafish", 
                                 "gsea", "gseaResults_kegg.rds"))
gseaResults_h<-readRDS(here("data", "zebrafish", 
                                 "gsea", "gseaResults_ire.rds"))
```



##  6. Results

- The combined p-values for each gene set in each comparison (contrast) 
can be seen below:

### 6.1. IRE gene sets

```{r}
gseaResults_ire$combTest
```

### 6.2. KEGG gene sets

```{r}
gseaResults_kegg$combTest
```

### 6.3. Hallmark gene sets

```{r}
gseaResults_h$combTest
```


## 7. Plots

### 6.1. Age-dependent effects of the fAD-like mutation

```{r fig.width=11}
# extract up/downregulated gene proportions from mroast
x_props <- gseaResults_h$indivTest$normoxia_6mth_mutant_vs_wt$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown, ) %>% 
  dplyr::mutate(contrast="normoxia_6mth_mutant_vs_wt",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

# extract proportions from mroast
y_props <- gseaResults_h$indivTest$normoxia_24mth_mutant_vs_wt$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="normoxia_24mth_mutant_vs_wt",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

# create the plot
# adapted from https://stackoverflow.com/questions/4559229/drawing-pyramid-plot-using-r-and-ggplot2
xy_plot <- bind_rows(x_props, y_props) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, pattern = "_", replacement = " "),
                Geneset = gsub(x = Geneset, pattern = "HALLMARK", replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset","NGenes","contrast")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, levels = c("notSig", "down", "up")),
                contrast = factor(contrast, levels = c("normoxia_6mth_mutant_vs_wt","normoxia_24mth_mutant_vs_wt"))) %>%
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col() +
  ggpol::facet_share(~contrast, scales = "free") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  coord_flip() +
  theme(axis.line.y = element_blank())

xy_plot
```


## Export Objects

- Gene set enrichment analysis results:

```{r eval=FALSE}
gseaResults_ire %>% saveRDS(here("data", "zebrafish", 
                                 "gsea", "gseaResults_ire.rds"))
gseaResults_kegg %>% saveRDS(here("data", "zebrafish", 
                                 "gsea", "gseaResults_kegg.rds"))
gseaResults_h %>% saveRDS(here("data", "zebrafish", 
                                 "gsea", "gseaResults_ire.rds"))
```

